Parsing Postfix log files

### 2008-04-03 doc/logparser.tex
Further corrections from Carl; the majority are now done.  The remainder are more difficult.

XXX PICK A NAME AND INTRODUCE IT\@.

@@

Most mail server administrators will have performed some basic processing
of the logs produced by their mail server at one time or another, whether
it was to debug a problem, explain to a user why their mail is being
rejected, or check whether new anti-spam measures are working.  The more
adventurous will have generated statistics to show how many hits each of
their anti-spam measures has gotten in the last week, and possibly even
generated some graphs to clearly illustrate the point to management or
users.\footnote{This was the author's first real foray into processing
Postfix logs.}  Very few will have performed in-depth parsing and analysis
of their logs, where the parsing must correlate the log lines
per-connection or per-queueid rather than processing log lines
independently.  One of the barriers to this type of processing is the
unstructured nature of Postfix logs, where each log line was added on an ad
hoc basis as a requirement was discovered or new functionality was
added.\footnote{A history of all changes made to Postfix is distributed
with the source code, available from \url{http://www.postfix.org/}} Further
complication arises because the set of rejection messages is not fixed: new
messages can be added by the administrator with custom checks; every
\RBL{}\footnote{This document is supplied with a glossary, see
appendix~\ref{Glossary}.} returns a different explanatory message; policy
servers may log different messages depending on the characteristics of the
connection; there are many ways in which the log lines may differ between
servers, even within the same organisation --- servers may be configured
differently, or running different version of Postfix.  This paper documents
the difficult process of parsing Postfix logs, presenting a program which
parses logs and places the resulting data into a database for later use.
The gathered data can then be used to optimise current anti-spam measures,
provide a baseline to test new anti-spam measures against, or to produce
statistics showing how effective those measures are.  There are numerous
other uses for such data: improving server performance by identifying
troublesome destinations and reconfiguring appropriately; identifying
regular high volume uses (e.g.\ customer newsletters) and restricting those
uses to off-peak times; detecting virus outbreaks which propagate via
email; as a base for billing customers on a shared server.  Preserving the
raw data enables users to develop a multitude of uses far beyond those
conceived of by the author.

@@

@@

Postfix's design follows the Unix philosophy of \textit{Write programs that
do one thing and do it well\/}~\cite{unix-philosophy},\footnote{This
principle was also followed when designing the parser (\sectionref{parser
design}), notably having separate actions rather than one huge action.} and
is separated into various component programs to perform the tasks required
of an \MTA{}\@: receive mail, send mail, local delivery of mail, etc. ---
full details can be found at~\cite{postfix-overview}.  Each log line
contains the name of the Postfix program which produced it, and this
information is used when determining which rules should be used to parse
each log line (see \sectionref{rule characteristics} for details.)  This
design has positive security implications also: those components which
interact with other hosts are not privileged, and those components which
are privileged do not interact with other hosts.

@@

    \item [sender] The sender's email address.  This can change during one
        single connection, when the connection is reused to send multiple
        mails.  XXX BETTER EXPLANATION\@.

@@

The most likely cause of mis-identification is if a mail generated
internally by Postfix is identified as a bounce notification but it is
actually a different type of message; arguably this is a benefit rather
than a drawback, as other mails generated internally by Postfix will be
handled correctly.

@@

It is important to compare and contrast newly developed programs,
algorithms and parsers against those already available, to accurately judge
what, if any, improvements are delivered by the newcomers.  There are not
that many previously developed Postfix log parsers, indeed it was quite
difficult to find ten parsers to review for this project, and the
functionality offered ranges from quite basic to much more mature,
depending on the needs of the creator.  The programs are not reviewed from
an independent viewpoint; the objective is to compare and contrast each
program with \parsername{}.  One important difference between \parsername{}
and the parsers reviewed is that only \parsername{} makes it possible to
parse new log lines without modifying the core parsing algorithm; see
\sectionref{why separate rules, actions and framework?} for a discussion of
why the separation of rules from the algorithm is beneficial.

@@

@@

        Sawmill is a general purpose commercial product which parses 687
        log file formats (correct as of 2007/12/04) and produces reports.
        Its data extraction facilities are quite limited, though it does
        extract three different sets of data for Postfix (one is beta as of
        2007/12/04), but they do not appear to be interlinked, nor does it
        save sufficient data for the purposes of this project.  No attempt
        is made to correlate log lines or deal with the difficulties
        documented in \sectionref{complications} \sectionref{additional
        complications}.\footnote{If any attempt is made there is no
        reference to it in the documentation available on the website.} The
        source code is available in an obfuscated form only (presumably for
        a fee), and the product is quite expensive, as it requires a
        license per report which is to be generated; in contrast
        \parsername{} is free and the code is freely available.  The web
        interface allows creation of dynamic reports based on any field,
        but due to the associated cost the author has not experimented with
        it.  Presumably, as the company charges per report the user wishes
        to generate, the data store (if there is one) is deliberately
        inaccessible and undocumented to prevent the user bypassing the
        program and generating their own reports.

        \url{http://www.thesawmill.co.uk/formats/postfix.html} \newline
        Fields extracted: from, to, server, UID, relay, status, number of
        recipients, origin hostname, origin \IP{} and virus.  The fields
        \textit{server}, \textit{uid\/} and \textit{virus\/} are not
        explained in their documentation: \textit{server\/} is probably the
        server the mail is delivered to, and \textit{uid\/} might be the
        uid of the user submitting mail locally.  Postfix does not perform
        any form of virus checking (though it has many options for
        cooperating with an external virus scanner), so the
        \textit{virus\/} field is a mystery.

@@

\glossary{name={cleanup},description={
    Cleanup processes all incoming mail, sitting between qmgr and the
    programs which accept mail (smtpd and pickup).  It removes duplicate
    recipient addresses, inserts missing headers, and optionally rewrites
    addresses if configured to do so.  The definitive
    documentation is \url{http://www.postfix.org/cleanup.8.html}.
}}

>>>>

XXX EITHER PICK A BETTER WORD OR GIVE A BETTER EXPLANATION OF EMERGENT
BEHAVIOUR --- SECTIONS 5.4.4 AND 8\@.

XXX FIGURE OUT THE CAUSE OF THE ABORTED DELIVERY ATTEMPTS IN SECTION
5.8.2\@.

XXX TEST OTHER PARSERS ON THE 93 TEST LOG FILES\@.

XXX POSSIBLE CONFERENCES\@: SIGAI IN BRITAIN, AND AICS'S AI-2008 (CAMBRIDGE
IN DECEMBER)\@.

@@

Most mail server administrators will have performed some basic processing
of the logs produced by their mail server at one time or another, whether
it was to debug a problem, explain to a user why their mail is being
rejected, or check whether new anti-spam measures are working.  The more
adventurous will have generated statistics to show how many hits each of
their anti-spam measures has gotten in the last week, and possibly even
generated some graphs to clearly illustrate the point to management or
users.\footnote{This was the author's first real foray into processing
Postfix logs.}  Very few will have performed in-depth parsing and analysis
of their logs, where the parsing must correlate the log lines
per-connection or per-queueid rather than processing log lines
independently.  One of the barriers to this type of processing is the
unstructured nature of Postfix logs, where each log line was added on an ad
hoc basis as a requirement was discovered or new functionality was
added.\footnote{A history of all changes made to Postfix is distributed
with the source code, available from \url{http://www.postfix.org/}} Further
complication arises because the set of rejection messages is not fixed: new
messages can be added by the administrator with custom checks; every
\RBL{}\footnote{This document is supplied with a glossary, see
appendix~\ref{Glossary}.} returns a different explanatory message; policy
servers may log different messages depending on the characteristics of the
connection; there are many ways in which the log lines may differ between
servers, even within the same organisation --- servers may be configured
differently, or running different version of Postfix.  This paper documents
the difficult process of parsing Postfix logs, presenting \PLP{}, a program
which parses logs and places the resulting data into a database for later
use.  The gathered data can then be used to optimise current anti-spam
measures, provide a baseline to test new anti-spam measures against, or to
produce statistics showing how effective those measures are.  There are
numerous other uses for such data: improving server performance by
identifying troublesome destinations and reconfiguring appropriately;
identifying regular high volume uses (e.g.\ customer newsletters) and
restricting those uses to off-peak times; detecting virus outbreaks which
propagate via email; as a base for billing customers on a shared server.
Preserving the raw data enables users to develop a multitude of uses far
beyond those conceived of by the author.

@@

\label{motivation}

@@

Postfix's design follows the Unix philosophy of \textit{Write programs that
do one thing and do it well\/}~\cite{unix-philosophy},\footnote{This
principle was also followed when designing the parser (\sectionref{parser
design}), notably having separate actions rather than one huge action.} and
is separated into various component programs to perform the tasks required
of an \MTA{}\@: receive mail, send mail, local delivery of mail, etc. ---
full details can be found at~\cite{postfix-overview}.  Each log line
contains the name of the Postfix program which produced it, and this
information is used when determining which rules should be used to parse
each log line (see \sectionref{rule characteristics} for details).  This
design has positive security implications also: those components which
interact with other hosts are not privileged,\footnote{Privilege in this
case means the power to perform actions that are limited to the
administrator, and not available to ordinary users.} so bugs in those
components will not give an attacker extra privileges; those components
which are privileged do not interact with other hosts, making it much more
difficult for an attacker to exploit any bugs which may exist in those
components.

@@

    \item [sender] The sender's email address.  Because multiple mails may
        be delivered during one connection, there may be different sender
        addresses in the results for one connection; however there should
        not be different sender addresses in the results for one email.

@@

If a mail is mis-classified as a bounce message it will almost certainly
have been generated internally by Postfix; arguably mis-classification in
this case is a benefit rather than a drawback, as other mails generated
internally by Postfix will be handled correctly.

@@

It is important to compare and contrast newly developed programs,
algorithms and parsers against those already available, to accurately judge
what, if any, improvements are delivered by the newcomers.  There are not
that many previously developed Postfix log parsers, indeed it was quite
difficult to find ten parsers to review for this project, and the
functionality offered ranges from quite basic to much more mature,
depending on the needs of the creator.  The programs are not reviewed from
an independent viewpoint; the objective is to compare and contrast each
program with \parsername{}.  There are some important differences between
\parsername{} and the parsers reviewed here:

    \item Only \parsername{} makes it possible to parse new log lines
        without modifying the core parsing algorithm; see \sectionref{why
        separate rules, actions and framework?} for a discussion of why the
        separation of rules from the algorithm is beneficial.

    \item \parsername{} does not provide any reporting based on the data it
        gathers.  This is by design: the parser is responsible for parsing
        logs and extracting data --- further processing of that data is a
        separate concern, deferred to another program.  Using an \SQL{}
        database simplifies the process of generating such reports; some
        simple examples are given in \sectionref{motivation}.

@@

In this section text in italics is a quote extracted from the appropriate
program's documentation.

@@

        Sawmill is a general purpose commercial product which parses 687
        log file formats (correct as of 2007/12/04) and produces reports.
        Its data extraction facilities are quite limited, though it does
        extract three different sets of data for Postfix (one is beta as of
        2007/12/04), but they do not appear to be interlinked, nor does it
        save sufficient data for the purposes of this project.\footnote{The
        data extracted by Sawmill is described later in this review, and
        the data extracted by \PLP{} is described in
        \sectionref{connections table} and \sectionref{results table}.}  No
        attempt is made to correlate log lines or deal with the
        difficulties documented in \sectionref{complications}
        \sectionref{additional complications}.\footnote{If any attempt is
        made there is no reference to it in the documentation available on
        the website.} The source code is available in an obfuscated form
        only (presumably for a fee), and the product is quite expensive, as
        it requires a license per report which is to be generated; in
        contrast \parsername{} is free and the code is freely available.
        The web interface allows creation of dynamic reports based on any
        field, but due to the associated cost the author has not
        experimented with it.  Presumably, as the company charges per
        report the user wishes to generate, the data store (if there is
        one) is deliberately inaccessible and undocumented to prevent the
        user bypassing the program and generating their own reports.

        \url{http://www.thesawmill.co.uk/formats/postfix.html} \newline
        Fields extracted: from, to, server, UID, relay, status, number of
        recipients, origin hostname, origin \IP{} and virus.  The fields
        \texttt{server}, \texttt{uid} and \texttt{virus} are not explained
        in their documentation: \texttt{server} is probably the server the
        mail is delivered to, and \texttt{uid} might be the uid of the user
        submitting mail locally.  Postfix does not perform any form of
        virus checking (though it has many options for cooperating with an
        external virus scanner), so the \texttt{virus} field is a mystery.

@@

\glossary{name={cleanup},description={
    Cleanup processes all incoming mail after it has been accepted and
    before it is delivered.  It removes duplicate recipient addresses,
    inserts missing headers, and optionally rewrites addresses if
    configured to do so.  The definitive documentation is
    \url{http://www.postfix.org/cleanup.8.html}.
}}
